Preregistration: Changes

### 2024-04-04 manuscript.Rmd
Substantial updates:  - Shortened manuscript  - Implemented peer-feedback

abstract          : "In times of crisis such as the COVID-19 pandemic, citizens need to stay informed about recent political events. To this end, people increasingly use social media. However, because social media are particularly engaging, many find it hard to disconnect, especially during times of crisis. Using data from the Austrian Corona Panel Project consisting of 3,485 participants from 34 waves, controlling for several stable and varying confounders, the results showed that COVID-19 related social media use did not meaningfully reduce well-being. Other factors such as health, income, exercise, or internal locus of control showed larger and meaningful effects."
keywords          : "COVID-19, well-being, social media, news use, panel study."

@@

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            :
  papaja::apa6_pdf:
    latex_engine  : xelatex
# output            : papaja::apa6_docx
# output            : html_document
---

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021].
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022].
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children.
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question
[e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021],
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

This study investigates how different _facets_ of well-being are affected by different _types_ and different _channels_ of communication [@meierComputermediatedCommunicationSocial2020a].
Building on the typology of subjective well-being [@dienerAdvancesOpenQuestions2018], three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used [@verduynPassiveFacebookUsage2015], I further distinguish three types of use and five popular channels. 
The types of use include reading, liking and sharing, and posting COVID-19 related content.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting and liking COVID-19 related news.
Liking and sharing are combined as they both represent low-threshold, platform-ingrained, easily quantifiable interactions.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are short-term, with well-being after some time returning to prior levels [@dienerAdvancesOpenQuestions2018].
Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
So although well-being can be affected by external events and factors, this does not happen easily.

Can social media use be such a factor?
Current literature overviews suggest that social media use on average does seem to decrease well-being [@meierComputermediatedCommunicationSocial2020a].
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small [@meierComputermediatedCommunicationSocial2020a].
These small findings can be explained with the differential susceptibility of media effects model [@valkenburgDifferentialSusceptibilityMedia2013], which states that there is substantial _variation_ of media effects for individual users.
Whereas for some users social media are more beneficial, for others they are more harmful.
On average, however, and this is central for this study here, effects tend to be small [@valkenburgDifferentialSusceptibilityMedia2013].
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral [@beyensSocialMediaUse2021].
Whether or not effects are positive or negative depend on (a) dispositional factors (e.g., personality, temperament, gender), (b) developmental factors (e.g., age, developmental tasks), (c) and social factors (e.g., environment, norms, upbringing).
Finally, effects depend also on the content that is consumed.
If the content is aligned with dispositions, developmental capacities, and converging contexts, effects tend to be stronger [@valkenburgDifferentialSusceptibilityMedia2013].

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

Second, while mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973].
Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.
If those beneficial media effects are missing, people will spend their time elsewhere.
And social media, in general, offer several beneficial effects, explaining why they are used that much.
They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].
In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory this indirectly suggest that average effects on well-being are likely not particularly negative.

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic [@twitterTalkingMentalHealthAwarenessTwitter2020]. 
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@atheyDigitalPublicHealth2023; @huntSocialMediaBased2022]. 
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022].
Social media platforms are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts. 
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@stainbackCOVID1924News2020; @yuePassiveSocialMedia2022]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, engaging more on social media was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
Several studies reported mostly neutral effects of social media use on well-being indicators [@edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

In conclusion, given these mixed empirical results, together with the observation that social media effects on well-being are very small in general, and that several plausible theoretical mechanisms exist for both positive and negative effects, I expect that COVID-19 related communication on social media should not be decidedly positive or negative.
It seems most likely that both positive and negative coexist, but that on average using social media for COVID-19 related reasons should not have substantial effects on well-being.

@@

Exploring the hypothesis entails establishing the criteria for discerning a 'trivial effect size.' 
To achieve this, it is necessary to define the smallest effect size of interest (SESOI) [@anvariUsingAnchorbasedMethods2021]. 
In this context, a trivial effect should fall below the threshold set by the SESOI (detailed below).
Determining what constitutes a minimally intriguing, nontrivial effect is a matter with normative implications, making it challenging to arrive at a definitive, singular consensus. 
Nonetheless, it remains both essential and beneficial to strive for a reasonable benchmark. 
With that in mind, I propose the following SESOI as a suitable reference point for this study:

> SESOI: If a heavy user of COVID-19 related social media news suddenly _stops_ using social media altogether, this should have a _noticeable_ impact on their overall well-being.

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
According to @normanInterpretationChangesHealthrelated2003, people can reliably notice seven levels of change in satisfaction with health.
So if satisfaction is measured on a 7-point scale, a four unit change in social media use should result in a one unit change in life satisfaction. 
In this study, life satisfaction was measured on an 11-point scale, and affect on a 5-point scale. 
Transposed to this scaling, the SESOI for life satisfaction is _b_ =  ±.30, and for positive and negative affect _b_ = ±.15 (for more information, see online supplementary material).

# Method
## Preregistration

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, _p_-value) were preregistered on the Open Science Framework, accessible here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyze data from an already existing large-scale data set, the preregistration was done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases, it was impossible to execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when seeing the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online [companion website](https://XMtRA.github.io/Austrian_Corona_Panel_Project) (https://XMtRA.github.io/Austrian_Corona_Panel_Project).

## Sample

The data come from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020], which is a large-scale standalone panel study. 
The data are hosted on AUSSDA, are publicly available here (https://doi.org/10.11587/28KQNS), and consist of 34 waves.
Participants were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Panel members were incentivized with 180 credit points for each wave of the study.
The study was conducted between March 2020 and February 2023.
Between March 2020 and July 2020, the intervals between waves were weekly, until May 2022 (wave 32) monthly, and afterward after 5 months.
Each wave consists of at least 1,500 respondents.
Panel mortality was compensated through a continuous re-acquisition of new participants. 
The sample size was _N_ = `r d_raw %>% nrow() %>% prettyNum(big.mark = ",")`, with overall `r d_long %>% nrow() %>% prettyNum(big.mark = ",")` observations.
For an overview of the study set-up, see Figure \@ref(fig:study-desc).

```{r study-desc, warning=F, message=F, fig.cap="Overview of study set-up", out.width = "1\\textwidth", warning=F, message=F, prompt=F}
knitr::include_graphics("figures/fig_study_description.png")
```

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age. 
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets). 
Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements. 
The participants provided their written informed consent to participate in this study.
The average age was `r mean(2021 - d_wide$year_birth, na.rm = T) %>% round(0)` years, `r mean(d_wide$male, na.rm = T) %>% round(., 2) * 100` percent were male, `r (d_wide %>% filter(.$edu_fac %in% c("State college", "Bachelor", "Master", "PhD"))) %>% nrow() %>% divide_by(sum(table(d_wide$edu_fac))) %>% multiply_by(100) %>% round(0)` percent had a University degree, and `r (table(d_wide$employment_fac)["Unemployed"] / sum(table(d_wide$employment_fac))) %>% unname() %>% multiply_by(100) %>% round(., digits = 0)` percent were currently unemployed.

@@

The hypothesis is analyzed using the interval testing approach as proposed by @dienesUsingBayesGet2014.
To illustrate, let us consider the case of life satisfaction [SESOI: ±.30]. 
Here, the null-region lies between -.30 and +.30.
If the 95% confidence interval falls completely within the null-region (e.g., _b_ = -.05, [95% CI: -.15, .05]), the hypothesis that the effect is trivial is supported.
If the confidence interval falls completely outside of the null-region (e.g., _b_ = -.40, [95% CI: -.45, -.35]), the hypothesis is rejected and the existence of a meaningful negative effect is supported.
If the confidence interval and the null region overlap (e.g., _b_ = -.30, [95% CI: -.35, -.25]), the hypothesis is not supported and the results are considered inconclusive, while a meaningful positive effect is rejected.
If the confidence interval exceeds both sides of the null region (e.g., _b_ = -.025, [95% CI: -.40, .35]), the hypothesis is not supported and judgement is suspended.
For an illustration, see Figure \@ref(fig:sesoi).

@@

When analyzing causality through longitudinal designs, it is important to address several critical aspects. An inherent requirement of analyzing causal effects within non-experimental designs is adopting an internal perspective, focusing on _within-person effects_ [@hamakerWhyResearchersShould2014]. This entails evaluating how alterations in an individual's media consumption directly impact their well-being. Comparisons between individuals, as observed in longitudinal data, lack the granularity needed to understand these causal dynamics [@hamakerWhyResearchersShould2014]. Consequently, this study exclusively investigates within-person effects to shed light on this intricate relationship.

>>>>

abstract          : "In times of crisis such as the COVID-19 pandemic, citizens need to stay informed about recent political events. To this end, people increasingly use social media. However, because social media are particularly engaging, many find it hard to disconnect, especially during times of crisis. In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. Using data from the Austrian Corona Panel Project consisting of 3,485 participants from 34 waves, this research question was analyzed using random effects within between models, controlling for several stable and varying confounders. The results showed that COVID-19 related social media use did not meaningfully reduce well-being. Other factors such as health, income, exercise, or internal locus of control showed larger and meaningful effects."
keywords          : "COVID-19, well-being, social media, news use, panel study."

@@

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
# output            :
#   papaja::apa6_pdf:
#     latex_engine  : xelatex
# output            : papaja::apa6_docx
output            : html_document
---

@@

# Research Transparency Statement

Conflicts of interest: The author declares no conflicts of interest. Funding: This research did not receive external funding. Artificial intelligence: Chat GPT was used for language polishing. Ethics: This research complies with the Declaration of Helsinki (2023), aside from the requirement to preregister human subjects research. Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements. The participants provided their written informed consent to participate in this study. Computational reproducibility: All analyses and the manuscript itself are completely reproducible (https://XMtRA.github.io/Austrian_Corona_Panel_Project).

Preregistration: The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, _p_-value) were preregistered on the Open Science Framework (https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3).
Because in this study data from an already existing data set was analyzed, the preregistration was done prior to accessing the data.
In some cases, it was impossible to execute the analyses as originally planned, for example because some properties of the variables only became apparent when seeing the actual data.
All changes are reported online (https://xmtra.github.io/Austrian_Corona_Panel_Project/preregistration_changes.html).
Materials: All study materials are publicly available (https://xmtra.github.io/Austrian_Corona_Panel_Project/items_english.html). Data: All primary data are publicly available (https://doi.org/10.11587/28KQNS). Analysis scripts: All analysis scripts are publicly available (https://XMtRA.github.io/Austrian_Corona_Panel_Project).

# Introduction

Amidst the COVID-19 pandemic, staying informed became paramount, prompting heavy reliance on social media for updates, with use being at an all time high [@statistaAverageDailyTime2021]. 
The phenomenon of "doomscrolling" emerged as individuals struggled to disengage from COVID-19-related news [@buchananBriefExposureSocial2021; @sharmaDarkEndTunnel2022], sparking concerns about its impact on mental health [@sandstromDoomscrollingCOVIDNews2021]. 
While initial research began exploring this question [e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021], the extent of social media's influence on well-being remains largely unknown. 
This study aims to assess the effects of different social media usage patterns on individual well-being using a comprehensive longitudinal dataset spanning 34 waves, offering insights into within-person causal relationships.

<!-- During the COVID-19 pandemic it was critical to stay informed regarding the latest developments. -->
<!-- How dangerous is the virus?  -->
<!-- In what region is it spreading?  -->
<!-- How is it transmitted?  -->
<!-- What are the current safety regulations? -->
<!-- To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021]. -->
<!-- Some actually could not stop using social media to learn about COVID-19 related news, and a new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022]. -->
<!-- Many users were glued to their screens and found it hard to pursue other relevant activities [@buchananBriefExposureSocial2021]. -->
<!-- In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021]. -->
<!-- Although research has begun addressing this question -->
<!-- [e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021], -->
<!-- it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being. -->
<!-- This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being. -->

@@

This study investigates how different _facets_ of subjective well-being are affected by different _types_ and different _channels_ of communication [@meierComputermediatedCommunicationSocial2020a].
Building on the typology of subjective well-being---defined as "referring to the various types of subjective evaluations of one’s life, including both cognitive evaluations and affective
feelings" [@dienerAdvancesOpenQuestions2018, p. 3]---three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.

Social media is a broad term, encompassing subdimensions such as social networking sites (e.g., Facebook or Instagram), instant messengers (e.g., WhatsApp or Signal), or interactive video-platforms (e.g., YouTube or TikTok) [@carrSocialMediaDefining2015].
Effects of social media depend on how they are used: while purposeful, active, social use is generally related to positive outcomes, passive, non-purposeful and asocial use is related to negative outcomes [@frisonExploringRelationshipsDifferent2015; @verduynSocialNetworkingSites2022].
In this study, I hence distinguish three types of use and five popular channels. 
The types of use include posting (active use), reading (passive use), and liking and sharing (low-threshold active use) COVID-19 related content.
<!-- In doing so, this study analyzes social media use focused on COVID-19 related content. -->
<!-- , which includes posting thoughts about the pandemic, reading posts and comments , or liking and sharing COVID-19 related content. -->
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

<!-- How easily can well-being be affected by external influences? -->
<!-- In general, according to the set-point theory, well-being is considered to be surprisingly stable [@lykkenHappinessWhatStudies1999]. -->
<!-- Although specific events such as marriage or salary increase can have significant impacts on well-being, effects are often only short-term, with well-being after some time returning to prior levels [@dienerAdvancesOpenQuestions2018]. -->
<!-- Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007]. -->
<!-- So although well-being can be affected by external events and factors, it seems this does not happen easily. -->

<!-- Can social media use be such a factor? -->
Current literature overviews suggest that social media use on average does seem to decrease well-being [@meierComputermediatedCommunicationSocial2020a].
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small [@meierComputermediatedCommunicationSocial2020a].
These small findings can be explained with the differential susceptibility of media effects model [@valkenburgDifferentialSusceptibilityMedia2013], which states that there is substantial _variation_ of media effects for individual users.
<!-- Whereas for some users social media are more beneficial, for others they are more harmful. -->
<!-- On average, however, effects tend to be small [@valkenburgDifferentialSusceptibilityMedia2013]. -->
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral [@beyensSocialMediaUse2021].
<!-- Whether or not effects are positive or negative depend on (a) dispositional factors (e.g., personality, temperament, gender), (b) developmental factors (e.g., age, developmental tasks), (c) and social factors (e.g., environment, norms, upbringing) [@valkenburgDifferentialSusceptibilityMedia2013]. -->
<!-- Effects depend also on the content that is consumed---if the content is aligned with dispositions, developmental capacities, and converging contexts, effects tend to be stronger [@valkenburgDifferentialSusceptibilityMedia2013]. -->

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
<!-- Use can be stimulating or overwhelming, relaxing or boring. -->
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

Second, while mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973].
Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.
<!-- If those beneficial media effects are missing, people will spend their time elsewhere. -->
And social media, in general, offer several beneficial effects, explaining their ubiquitous use.
They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].

Third, and closely related, media use affords both both positive and negative mechanisms, which differently impact well-being outcomes.
These include factors like social comparison, inspiration, social connectedness, social support, displacement of in-person activities, information management, information overload, misinformation, reputation building, romantic connection, coping resources, or emotional contagion [@clarkSocialNetworkSites2018; @hallSocialMediaUse2022; @pelletierOneSizeDoesn2020; @vandenabeeleDigitalWellbeingDynamic2021; @verduynSocialNetworkingSites2022].
Given these differential mechanisms, strong one-sided effects appear less likely than nuanced and moderate effects.

<!-- In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to mood management theory, uses and gratifications theory, and multiple differential mechanisms, average effects on well-being are likely not particularly negative. -->

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
<!-- Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic. -->
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
<!-- Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals. -->
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic. 
<!-- Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times. -->
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@huntSocialMediaBased2022], which ultimately benefit well-being.
<!-- Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility. -->

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
<!-- Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users. -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
<!-- The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022]. -->
Discussions around COVID-19 are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
<!-- Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.  -->
<!-- Such experiences threaten mental well-being and might contribute to feelings of distress and isolation. -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].

<!-- There is still limited empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically. -->
Echoing the theoretical rationales outlines above, empirical studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 content led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@stainbackCOVID1924News2020]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Doom-scrolling was associated with negative emotional experiences [@buchananBriefExposureSocial2021].
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, identification with social media networks was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
Several studies reported mostly neutral or dual effects of social media use on well-being indicators [@edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
<!-- Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being. -->

@@

In conclusion, given these mixed empirical results, together with the observation that social media effects on well-being are very small in general, and that several plausible theoretical mechanisms exist for both positive and negative effects, I expect that COVID-19 related communication on social media should not be decidedly positive or negative.
<!-- It seems most likely that both positive and negative coexist, but that on average using social media for COVID-19 related reasons should not have substantial effects on well-being. -->

@@

Exploring the hypothesis entails establishing the criteria for discerning a 'trivial effect size.' 
To achieve this, it is necessary to define the smallest effect size of interest (SESOI) [@anvariUsingAnchorbasedMethods2021].
<!-- , which helps determine an effect's practical significance.  -->
In this context, a trivial effect should fall below the threshold set by the SESOI (detailed below).
Determining what constitutes a minimally intriguing, nontrivial effect is a matter with normative implications, making it challenging to arrive at a definitive, singular consensus. 
<!-- Nonetheless, it remains both essential and beneficial to strive for a reasonable benchmark.  -->
I propose the following SESOI as a suitable reference point for this study:

> SESOI: If a heavy user of COVID-19 related social media content suddenly _stops_ using social media altogether, this should have a _noticeable_ impact on their overall well-being.

<!-- What does this mean practically and how can it be operationalized? -->
Put concretely, in this study COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
According to @normanInterpretationChangesHealthrelated2003, people can reliably notice seven levels of change in satisfaction with health.
So if satisfaction is measured on a 7-point scale, a four unit change in social media use should result in a one unit change in life satisfaction. 
Life satisfaction was measured on an 11-point scale, and affect on a 5-point scale. 
Transposed to this scaling, the SESOI for life satisfaction is _b_ =  ±.30, and for positive and negative affect _b_ = ±.15 (see online supplementary material).

# Method
## Sample

The data come from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020], which is a large-scale standalone panel study, consisting of 34 waves.
<!-- The data are hosted on AUSSDA, are publicly available here (https://doi.org/10.11587/28KQNS). -->
<!-- Participants were sampled from a pre-existing online access panel provided by the company Marketagent, Austria. -->
<!-- Panel members were incentivized with 180 credit points for each wave of the study. -->
The study was conducted between March 2020 and February 2023.
Between March 2020 and July 2020, the intervals between waves were weekly, until May 2022 (wave 32) monthly, and afterward after 5 months.
Each wave consists of at least 1,500 respondents.
Panel mortality (on average ca. 5 percent per wave) was compensated through a continuous re-acquisition of new participants (on average ca. 80 new respondents per wave). 
The sample size was _N_ = `r d_raw %>% nrow() %>% prettyNum(big.mark = ",")`, with overall `r d_long %>% nrow() %>% prettyNum(big.mark = ",")` observations.

<!-- For an overview of the study set-up, see Figure \@ref(fig:study-desc). -->

<!-- ```{r study-desc, warning=F, message=F, fig.cap="Overview of study set-up", out.width = "1\\textwidth", warning=F, message=F, prompt=F} -->
<!-- knitr::include_graphics("figures/fig_study_description.png") -->
<!-- ``` -->

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age. 
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets). 
The average age was `r mean(2021 - d_wide$year_birth, na.rm = T) %>% round(0)` years, `r mean(d_wide$male, na.rm = T) %>% round(., 2) * 100` percent were male, `r (d_wide %>% filter(.$edu_fac %in% c("State college", "Bachelor", "Master", "PhD"))) %>% nrow() %>% divide_by(sum(table(d_wide$edu_fac))) %>% multiply_by(100) %>% round(0)` percent had a University degree, and `r (table(d_wide$employment_fac)["Unemployed"] / sum(table(d_wide$employment_fac))) %>% unname() %>% multiply_by(100) %>% round(., digits = 0)` percent were currently unemployed.

@@

The hypothesis is analyzed using the interval testing approach as proposed by @dienesUsingBayesGet2014.
To illustrate, let us consider the case of life satisfaction [SESOI: ±.30]. 
Here, the null-region lies between -.30 and +.30.
<!-- If the 95% confidence interval falls completely within the null-region (e.g., _b_ = -.05, [95% CI: -.15, .05]), the hypothesis that the effect is trivial is supported. -->
If the confidence interval falls completely outside of the null-region (e.g., _b_ = -.40, [95% CI: -.45, -.35]), the hypothesis is rejected and the existence of a meaningful negative effect is supported.
<!-- If the confidence interval and the null region overlap (e.g., _b_ = -.30, [95% CI: -.35, -.25]), the hypothesis is not supported and the results are considered inconclusive, while a meaningful positive effect is rejected. -->
<!-- If the confidence interval exceeds both sides of the null region (e.g., _b_ = -.025, [95% CI: -.40, .35]), the hypothesis is not supported and judgement is suspended. -->
For an illustration, see Figure \@ref(fig:sesoi).

@@

<!-- When analyzing causality through longitudinal designs, it is important to address several critical aspects.  -->
Analyzing causal effects within non-experimental designs requires adopting an internal perspective, focusing on _within-person effects_ [@hamakerWhyResearchersShould2014]. 
This entails evaluating how alterations in an individual's media consumption directly impact changes in their own well-being. 
<!-- Comparisons between individuals, as observed in longitudinal data, lack the granularity needed to understand these causal dynamics [@hamakerWhyResearchersShould2014].  -->
Consequently, this study exclusively investigates within-person effects.
<!-- to shed light on this intricate relationship. -->
